### 2010-03-18 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: uniformize EnumT definitions
Desc@ICFP: {Free} is exponent, instead of subscript
Desc@ICFP: Fin specified in Agda-like notation
Desc@ICFP: correct presentation of substitution for Hutton's razor. Introduce spurious alignment issues.

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DId            & \Desc                                   \\
    \DUnit          & \Desc                                   \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{X}          &  X                                           \\
\descop{\DUnit}{X}        &  \Unit                                       \\
\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
\end{array}
}
\end{array}
\]

@@

\[\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \ldots          & \ldots
\end{array} \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\ldots                        &  \ldots
\end{array}
}\]

@@

\[
\begin{array}{ll}
\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
&
\stk{
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
\end{array}
}
\end{array}
\]

@@

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} ]) \\
                              & \left[\begin{array}{l}
                                      \DConst{I}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAlwaysD}}
\newcommand{\HExprID}{\C{ExprIndexedD}}
\newcommand{\HExprVarD}{\C{ExprD}_{\C{Var}}}
\newcommand{\HExprFreeD}{\C{ExprD}_{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}_{\C{Free}}}

@@

\[
\Ty \mapsto \EnumT{[\Enat \quad \Ebool]}
\]

@@

\[\stkc{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\begin{array}{ll}
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \left[
                 \begin{array}{l}
                   \EnumT{[\Eval{}  \Econd{}{}{}]} \red{,} \\
                   \LAM{ty}
                   \left[ %%! Should be read
                   \begin{array}{l}
                   \DConst{(\Val{ty})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
                   \end{array}
                   \right]
                 \end{array}
                 \right] %%! Should be \red
\\
\\
} &
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \left[
                 \begin{array}{l}
                   \LAM{ty} \case{ty}{\Enat \To \EnumT{[\Eplus{}{}]} \\ \Ebool \To \EnumT{[\Ele{}{}]}} \red{,} \\
                   \LAM{ty}
                   \case{ty}{
                     \Enat \To \DProd{\DVar{nat}}{\DVar{nat}} \\
                     \Ebool \To \DProd{\DVar{nat}}{\DVar{nat}}
                   }
                   \end{array}
                   \right]
}
\end{array}
}\]

@@

Consequently, replacing $\Val{ty}$ by $(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in
Figure~\ref{fig:hexpr-full} turns the language of closed term into a
language of opened terms with constants. \note{Is it polite and
  comprehensible to ask that?} This defines a new indexed description,
called $\HExprVarD$.

@@

\[\stk{
\begin{array}{@{}ll}
\discharge : & \PITEL{\V{ty}}{\Ty}
               \PITEL{\V{dom}}{\EnumU} \\
             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
             & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
               \IMu{\Ty}{\toIDesc{\HExprD}}{ty} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}{\gamma}{v}
\end{array}
}\]

@@

\newcommand{\substH}{\F{substExpr}}

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{ty}}{\Ty}
            \PITEL{\V{dom}}{\EnumU} 
            \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD}}{ty})}} \\
          &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
                             \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}} \\
          & \IMu{\Ty}{\toIDesc{\HExprVarD}}{\V{ty}} \To
            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic \bind. This is not surprising as we are defining a first-order
syntax with variables: our data-type enjoys more structure than what
we are given. In particular, this is reminiscent to a free monad,
where $\Eval{}$ is the \return introducing variables. Instead of
implementing $\substH$ in this special case, we shall implement it in
a generic setting, and express our example in that setting. However,
before entering the realm of generic structures, let us further
explore the universe of indexed descriptions.

@@

\[
\Rule{\Bhab{n}{\Nat}}
     {\Bhab{\FinZero}{\Fin{(\NatSuc{n})}}}
\qquad
\Rule{\Bhab{n}{\Nat} \quad 
      \Bhab{i}{\Fin{n}}}
     {\Bhab{\FinSuc{}}{\Fin{(\NatSuc{n})}}}
\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: n = \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
                        & \left[
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
                          \end{array}
                          \right]
\end{array}
}\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: 0            & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT [ \FinZero , \FinSuc{} ]) \\
                     &   &             & \left[
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{n}
                                         \end{array}
                                         \right]
\end{array}
}\]

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
identified the variable substitution operation as the \bind of a free
monad. The definition of $\substI$ above confirm our intuition: using
$\substI$, we should easily obtain $\substH$. To do so, we first have
to massage the definition of our data-type, to exhibit its monadic
structure.

@@

The resulting data-type is called $\HExprFreeD$. By a simple unfolding
of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
corresponds to the syntax of closed terms, $\HExprD$. Similarly,
$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
corresponds to expressions with variables, $\HExprVarD$.

@@

\[\stk{
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{ty}}{\Ty}
             \PITEL{\V{dom}}{\EnumU} 
             \PITEL{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{ty})}} \\
           &  \PI{\V{\sigma}}{\PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom}{ty}) \To
                              \IMu{\Ty}{(\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}})}{ty}} \\
           & \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{} + \Var{dom}{}}}}{\V{ty}} \To
             \IMu{\Ty}{\toIDesc{\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}}}{\V{ty}} \\
\end{array} \\
\substH\: ty\: dom\: \gamma\: \sigma\: term \mapsto 
    \substI\: \Ty\: 
              (\Val{} + \Var{dom}{})\: 
              \Val{}\:
              \HExprFreeD\:
              \sigma\:
              ty\:
              term
}\]

\note{Not using $\discharge$, that's a shame.}

@@

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT [ \DConst{},
                                              \DProd{}{},
                                              \DSigma{}{}, 
                                              \DPi{}{} ]) \\
                                  & \left[\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}\right]
\end{array}
}\]

>>>>

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DId            & \Desc                                   \\
    \DUnit          & \Desc                                   \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\Desc} \Desc         \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc 
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DId}{X}          &  X                                           \\
\descop{\DUnit}{X}        &  \Unit                                       \\
\descop{\DProd{D}{D'}}{X} &  \TIMES{\descop{D}{X}}{\descop{D'}{X}}       \\
\descop{\DSigma{S}{D}}{X} &  \SIGMA{\V{s}}{S} \descop{D\: s}{X}                \\
\descop{\DPi{S}{D}}{X}    &  \PI{\V{s}}{S} \descop{D\: s}{X}            
\end{array}
}
\end{array}
\]

@@

\[\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \ldots          & \ldots
\end{array} \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\ldots                        &  \ldots
\end{array}
}\]

@@

\[
\begin{array}{ll}
\stk{
\data \IDesc{} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
&
\stk{
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
\end{array}
}
\end{array}
\]

@@

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
}\]

@@

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT{[ \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} ]}) \\
                              & \left[\begin{array}{l}
                                      \DConst{I}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             \right]
\end{array}
}\]

@@

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAlwaysD}}
\newcommand{\HExprID}{\C{ExprIndexedD}}
\newcommand{\HExprVarD}[1]{\C{ExprD}_{\C{Var},#1}}
\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}^{\C{Free}}}

@@

\[
\Ty \mapsto \EnumT{[\Enat, \Ebool]}
\]

@@

\[\stkc{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\begin{array}{ll}
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \left[
                 \begin{array}{l}
                   \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
                   \LAM{ty}
                   \left[ %%! Should be read
                   \begin{array}{l}
                   \DConst{(\Val{ty})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{ty}}{\DVar{ty}}}
                   \end{array}
                   \right]
                 \end{array}
                 \right] %%! Should be \red
\\
\\
} &
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \left[
                 \begin{array}{l}
                   \LAM{ty} \case{ty}{\Enat \To \EnumT{[\Eplus{}{}]} \\ \Ebool \To \EnumT{[\Ele{}{}]}} \red{,} \\
                   \LAM{ty}
                   \case{ty}{
                     \Enat \To \DProd{\DVar{nat}}{\DVar{nat}} \\
                     \Ebool \To \DProd{\DVar{nat}}{\DVar{nat}}
                   }
                   \end{array}
                   \right]
}
\end{array}
}\]

@@

Consequently, replacing $\Val{ty}$ by $(\SUM{\Val{ty}}{\Var{dom}{ty}})$ in
Figure~\ref{fig:hexpr-full} turns the language of closed term into a
language of opened terms with constants. \note{Is it polite and
  comprehensible to ask that?} This defines a new indexed description,
called $\HExprVarD{dom}$.

@@

\[\stk{
\begin{array}{@{}ll}
\discharge : & \PITEL{\V{ty}}{\Ty}
               \PITEL{\V{dom}}{\EnumU} \\
             & \PI{\V{\gamma}}{\spi{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}} \\
             & (\SUM{\Val{ty}}{\Var{dom}{ty}}) \To
               \IMu{\Ty}{\toIDesc{\HExprD}}{ty} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: ty\: dom\: \gamma\: (\SumLeft x)  & \Con{\pair{\Eval{}}{x}{}} \\
\discharge\: ty\: dom\: \gamma\: (\SumRight v) & \switch{dom}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{dom}}}{ty})}{\gamma}{v}
\end{array}
}\]

@@

\newcommand{\substH}{\F{substExpr}}
\newcommand{\domNat}{dom_{\CN{nat}}}
\newcommand{\domBool}{dom_{\CN{bool}}}
\newcommand{\gammaNat}{\gamma_{\CN{nat}}}
\newcommand{\gammaBool}{\gamma_{\CN{bool}}}

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{\domNat}}{\EnumU} 
             \PITEL{\V{\gammaNat}}{\spi{\domNat}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domNat}}}{\Enat})}} \\
           & \PITEL{\V{\domBool}}{\EnumU} 
             \PITEL{\V{\gammaBool}}{\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
          &  \begin{array}{@{}ll}
             \PI{\V{\sigma}}{& \PITEL{\V{\domNat}}{\EnumU} \\
                             & \PITEL{\V{\gammaNat}}
                                     {\spi{\domNat}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domNat}}}{\Enat})}} \\
                             & \PITEL{\V{\domBool}}{\EnumU} \\
                             & \PITEL{\V{\gammaBool}}
                                     {\spi{\domBool}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\domBool}}}{\Ebool})}} \\
                             & \PI{\V{ty}}{\Ty} (\Val{ty} + \Var{dom_{ty}}{ty}) \To
                               \IMu{\Ty}{(\toIDesc{\HExprD})}{ty}}
             \end{array}\\
          & \PI{\V{ty}}{\Ty}
            \IMu{\Ty}{\toIDesc{\HExprVarD{dom_{ty}}}}{\V{ty}} \To
            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

Where $dom_{ty}$ is short for
 $\case{ty}{\Enat  \To \domNat
         \\ \Ebool \To \domBool}$

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic \bind. This is not surprising as we are defining a first-order
syntax with variables: our data-type enjoys more structure than what
we are given. In particular, this is reminiscent to a free monad,
where $\Eval{}$ is the \return\ introducing variables. The substitution
$\sigma$ is implemented from $\discharge$, by picking the domain and
context corresponding to the type:

\[
\sigma\: \domNat\: \gammaNat\: \domBool\: \gammaBool\: ty\: var \mapsto 
    \discharge\: ty\: dom_{ty}\: \gamma_{ty}\: var 
\]

Instead of implementing $\substH$ in this special case, we shall
implement it in a generic setting, and express our example in that
setting. However, before entering the realm of generic structures, let
us further explore the universe of indexed descriptions.

@@

\[
\stk{
\data \Fin{} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_n      & \Fin{(\NatSuc{n})}   \\
    \FinSuc{}_n     & \Fin{n} \To \Fin{(\NatSuc{n})}
\end{array}
}
\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: n = \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                        & \left[
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{m} \DConst{(n \PropEq \NatSuc{m})}} \\
                            \DSigma{\Nat}{\LAM{m} \DProd{\DVar{m}}{\DConst{(n \PropEq \NatSuc{m})}}}
                          \end{array}
                          \right]
\end{array}
}\]

@@

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: 0            & = & \multicolumn{2}{l}{\DSigma{\Void}{\Void} } \\
\FinD\: (\NatSuc{n}) & = & \DSigma{}{} & (\EnumT{[ \FinZero , \FinSuc{} ]}) \\
                     &   &             & \left[
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{n}
                                         \end{array}
                                         \right]
\end{array}
}\]

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD{dom}$ enjoy a monadic structure. We had
identified the variable substitution operation as the \bind of a free
monad. The definition of $\substI$ above confirm our intuition: using
$\substI$, we should easily obtain $\substH$. To do so, we first have
to massage the definition of our data-type, to exhibit its monadic
structure.

@@

The resulting data-type is called $\HExprFreeD$. By a simple unfolding
of definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\Val{}}$
corresponds to the syntax of closed terms, $\HExprD$. Similarly,
$$\FreeIMonad{\HExprFreeD}{\Ty}{(\LAM{ty} \Val{ty} + \Var{dom}{ty})}$$
corresponds to expressions with variables, $\HExprVarD{dom}$.

@@

\[\stk{
\substH\: \domNat\: \domBool\: 
            \gammaNat\: \gammaBool\:
            \sigma\: 
            ty\: 
            term \mapsto  \\
\;\;\ \begin{array}{ll}
       \substI\: & \Ty\: 
                  (\Val{} + \Var{dom_{ty}}{})\: 
                  \Val{}\:
                  \HExprFreeD\:  \\
                &
                  (\sigma\: \domNat\: \domBool\: \gammaNat\: \gammaBool)\:
                  ty\:
                  term
      \end{array}
}\]

@@

\[\stk{
\IDescFreeD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescFreeD~I \mapsto \DSigma{}{} & (\EnumT{[ \DConst{},
                                              \DProd{}{},
                                              \DSigma{}{}, 
                                              \DPi{}{} ]}) \\
                                  & \left[\begin{array}{l}
                                        \DConst{\Set}               \\
                                        \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                        \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                    \end{array}\right]
\end{array}
}\]
